---
title: "Cannabis Use in Canada"
author: "John W"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

The purpose of this notebook is to analyze the trends in the daily number of cannabis users in Canada for a period of 10 years, before legalization occurred in 2018. The dataset used by this notebook is available [here.](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3610059701)

# Analysis

## How has the amount of daily cannabis use in Canada changed over time?

```{r}
library(tidyverse)
library(janitor)
library(viridis)
library(scales)

data <- read.csv("36100597.csv") |>
  clean_names() |>
  filter(
    frequency_of_use != "Total, all frequencies",
    age_group != "15 years and over",
    age_group != "65 years and over",
    ref_date >= 2000
  )

df_plot <- data |>
  group_by(ref_date, age_group, frequency_of_use) |>
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop")

daily_data <- df_plot |>
  filter(frequency_of_use == "Daily")

ggplot(daily_data, aes(x = ref_date, y = value, fill = age_group)) +
  geom_col() +
  labs(
    title = "Daily Cannabis Use Over Time",
    subtitle = "Canada",
    x = "Year",
    y = "Number of Individuals",
    fill = "Age Group"
  ) +
  scale_y_continuous(labels = comma) +
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The plot shows daily cannabis use in Canada over time (2000-2018), broken down by age group in a stacked bar chart. The overall trend shows a steady and substantial increase in total daily users across the entire period. The total rises from roughly 600,000 individuals in 2000 to over 1.1 million by the late 2010s. The 25-44 age group is the largest contributor throughout the observation period, showing strong, consistent growth and driving much of the overall increase. The 45-64 segment is the fastest growing proportionally, starting relatively small but increasing sharply, especially after 2010. The 18-24 age group remains relatively stable compared to other groups, with some slight growth that levels off early on. The smallest group across all years is the 15-17 age group, which also sees some slight growth early on, then remains
relatively flat.

Growth in daily cannabis users accelerates after around 2010, with all adult age groups increasing. The increase is driven mainly by older age groups (25-44 and 45-64) rather than youth. The composition of users shifts over time toward older populations. Daily cannabis use becomes more common and widespread over time, with a clear shift toward increased use among adults, particularly middle-aged individuals.

# Conclusion

Daily cannabis use in Canada shows a clear, upward trend from 2000 to 2018, nearly doubling overall. Growth is primarily driven by adults, especially in the 25-44 age group. The 45-64 group shows the fastest relative increase, indicating rising use among older individuals. Younger groups remain relatively stable, contributing less to overall growth. A noticeable acceleration occurs after 2010, with increases across all adult groups. The overall composition shifts toward an older user population over time. Overall, daily cannabis use became more widespread and increasingly concentrated among middle-aged and older adults, rather than youth.