---
title: "Cannabis Use in Canada"
author: "John W"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

The purpose of this notebook is to analyze the demographics of daily cannabis use in Canada, leading up to legalization (which occurred in 2018). The data set used by this notebook is available  [here.](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3610059701)

# Analysis

## What is the distribution of daily cannabis use by age group?

```{r}
library(tidyverse)
library(janitor)
library(viridis)
library(scales)

data <- read.csv("36100597.csv") |>
  clean_names() |>
  filter(
    frequency_of_use != "Total, all frequencies",
    age_group != "15 years and over",
    age_group != "65 years and over",
    ref_date >= 2000
  )

daily_data <- data |>
  filter(frequency_of_use == "Daily") |>
  group_by(ref_date, age_group) |>
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") |>
  group_by(ref_date) |>
  mutate(prop = value / sum(value))

ggplot(daily_data, aes(x = ref_date, y = prop, fill = age_group)) +
  geom_area(alpha = 0.9) +
  labs(
    title = "Distribution of Daily Cannabis Use by Age Group Over Time",
    subtitle = "Canada (2000+)",
    x = "Year",
    y = "Proportion of Users",
    fill = "Age Group"
  ) +
  scale_y_continuous(labels = percent) +
  scale_fill_viridis_d() +
  theme_minimal()
```

This stacked area chart shows how the composition of daily cannabis users by age group in Canada has shifted since 2000. A few clear patterns stand out. The share of users in older groups, especially 45 to 64 years, has steadily increased over time. This group roughly doubles its proportion across the period from 2000 to 2018. In contrast, the 18 to 24 age group shows a consistent decline in share, indicating that daily use is becoming less concentrated among younger adults. The 25 to 44 age group remains the largest segment throughout, though its share gradually decreases as there is growth in older groups. The 15 to 17 age group is consistently the smallest and trends slightly downward over time. Daily cannabis is becoming more evenly distributed across older age groups with time, with a clear shift away from younger users toward middle-aged and older adults.

# Conclusion

Overall, the data shows a clear shift in daily cannabis use toward older age groups over time. While individuals aged 25 to 44 remain the largest group of users, their share is gradually declining as the proportion of users aged 45 to 64 increases significantly. At the same time, younger age groups—especially those aged 18 to 24—are making up a smaller share of daily users, with teens remaining a consistently minimal portion. Taken together, this indicates that daily cannabis use in Canada is becoming increasingly concentrated among middle-aged and older adults rather than younger populations.