---
title: "Cannabis Use in Canada"
author: "John W"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

The purpose of this notebook is to analyze the demographics of daily cannabis use in Canada, leading up to legalization (which occurred in 2018). The data set used by this notebook is available  [here.](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3610059701)

# Analysis

## How does daily cannabis use vary across age groups from 2000-2018?

```{r}
library(tidyverse)
library(janitor)
library(viridis)
library(scales)

data <- read.csv("36100597.csv") |>
  clean_names() |>
  filter(
    frequency_of_use != "Total, all frequencies",
    age_group != "15 years and over",
    age_group != "65 years and over",
    ref_date >= 2000
  )

daily_data <- data |>
  filter(frequency_of_use == "Daily") |>
  group_by(ref_date, age_group) |>
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop")

ggplot(daily_data, aes(x = age_group, y = value, fill = age_group)) +
  geom_boxplot() +
  labs(
    title = "Variation in Daily Cannabis Use Across Age Groups",
    subtitle = "Canada (2000-2018)",
    x = "Age Group",
    y = "Number of Users",
    fill = "Age Group"
  ) +
  scale_y_continuous(labels = comma) +
  scale_fill_viridis_d() +
  theme_minimal()
```

This boxplot illustrates how the number of daily cannabis users in Canada varied across age groups between 2000 and 2018, revealing clear differences in both typical usage and variability. The 15–17 age group has the lowest median—well under 100,000—and the narrowest interquartile range, indicating consistently low and stable usage over time. In contrast, the 18–24 group shows a much higher median, close to 300,000, while still maintaining a relatively tight interquartile range; however, the presence of several outliers suggests occasional spikes in usage. The 25–44 age group has the highest median, just under 300,000, along with a wide interquartile range and an extended upper whisker, reflecting both high typical use and notable variability. Meanwhile, the 45–64 group has a slightly lower median of around 200,000 but exhibits the greatest overall spread, with the widest interquartile range and long whiskers on both ends, indicating substantial fluctuations in daily cannabis use within this age group.

# Conclusion

Overall, the boxplot highlights clear differences in daily cannabis use across age groups in Canada from 2000 to 2018. Usage tends to increase from youth into early and middle adulthood, peaking in the 25–44 age group, before declining slightly in the 45–64 group. Younger individuals (15–17) show consistently low and stable usage, while older groups exhibit greater variability, particularly among those aged 45–64. These patterns suggest that cannabis use is not only more prevalent but also more variable in adult populations compared to younger age groups.