---
title: "Cannabis Use in Canada"
author: "John W"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

The purpose of this notebook is to analyze the demographics of daily cannabis use in Canada, leading up to legalization (which occurred in 2018). The data set used by this notebook is available  [here.](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3610059701)

# Analysis

## What is the distribution of daily cannabis use by age group?

```{r}
library(tidyverse)
library(janitor)
library(viridis)
library(scales)

data <- read.csv("36100597.csv") |>
  clean_names() |>
  filter(
    frequency_of_use != "Total, all frequencies",
    age_group != "15 years and over",
    age_group != "65 years and over",
    ref_date >= 2000
  )

df_plot <- data |>
  group_by(ref_date, age_group, frequency_of_use) |>
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop")

daily_data <- df_plot |>
  filter(frequency_of_use == "Daily")

ggplot(daily_data, aes(x = age_group, y = value, fill = age_group)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Daily Cannabis Use by Age Group",
    subtitle = "Canada (2000+)",
    x = "Age Group",
    y = "Number of Individuals",
    fill = "Age Group"
  ) +
  scale_y_continuous(labels = comma) +
  scale_fill_viridis_d() +
  theme_minimal()
```

The distribution of daily cannabis use in Canada since 2000 varies noticeably across age groups. Individuals aged 25–44 show the highest median number of users and the greatest variability, indicating both consistently high usage and significant year-to-year fluctuations. The 18–24 age group has a similarly high median but a much tighter distribution, suggesting more stable usage over time with fewer extreme changes. In contrast, the 45–64 group has a lower median and a wider spread, reflecting moderate usage with considerable variability across years. The 15–17 group consistently reports the lowest numbers and minimal variation, indicating stable and comparatively low usage. Overall, usage rises from adolescence to early middle age before declining in older groups, with variability generally increasing in the older age categories.

# Conclusion

Daily cannabis use in Canada is most concentrated among adults in the 25–44 age group, who not only have the highest number of users but also experience the greatest fluctuations over time. Younger adults aged 18–24 also contribute substantially, though their usage remains more consistent. In contrast, both adolescents and older adults show lower overall participation, with teens displaying particularly stable and minimal use. These patterns suggest that cannabis use is most prominent and dynamic during early to middle adulthood, while remaining relatively limited and stable at the younger and older ends of the population.