---
title: "Cannabis Use in Canada"
author: "John W"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

The purpose of this notebook is to analyze the demographics of daily cannabis use in Canada, leading up to legalization (which occurred in 2018). The data set used by this notebook is available  [here.](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3610059701)

# Analysis

## How does daily cannabis use vary across age groups from 2000-2018?

```{r}
library(tidyverse)
library(janitor)
library(viridis)
library(scales)

data <- read.csv("36100597.csv") |>
  clean_names() |>
  filter(
    frequency_of_use != "Total, all frequencies",
    age_group != "15 years and over",
    age_group != "65 years and over",
    ref_date >= 2000
  )

daily_data <- data |>
  filter(frequency_of_use == "Daily") |>
  group_by(ref_date, age_group) |>
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop")

ggplot(daily_data, aes(x = age_group, y = value, fill = age_group)) +
  geom_boxplot() +
  labs(
    title = "Variation in Daily Cannabis Use Across Age Groups",
    subtitle = "Canada (2000-2018)",
    x = "Age Group",
    y = "Number of Users",
    fill = "Age Group"
  ) +
  scale_y_continuous(labels = comma) +
  scale_fill_viridis_d() +
  theme_minimal()
```

This boxplot illustrates how daily cannabis use varies across age groups in Canada from 2000–2018. The 15–17 age group consistently shows the lowest median, at under 100,000 users, along with the smallest interquartile range, indicating relatively low and stable usage over time. In contrast, the 18–24 age group has a much higher median, close to 300,000, and includes several outliers, suggesting occasional years of unusually high or low usage. The 25–44 age group records the highest median, also near 300,000, but displays a wider interquartile range and a right-skewed distribution with a long upper whisker, indicating that while usage is generally high, it occasionally spikes substantially. The 45–64 age group shows a lower median, around 200,000, but the greatest overall variability, reflected in its wide interquartile range and extended whiskers on both ends, suggesting more fluctuation in usage levels over time compared to the other groups.

# Conclusion

The boxplot shows clear differences in daily cannabis use across age groups over the 2000–2018 period. Usage is lowest and most stable among those aged 15–17, while all adult age groups show substantially higher levels and greater variability. The 18–24 and 25–44 groups stand out with the highest medians, indicating that young and mid-adults are the primary drivers of daily use. However, the 25–44 and 45–64 groups also display wider spreads and more pronounced skewness, suggesting that usage in these populations is less consistent over time and subject to larger fluctuations. Overall, the data indicates that daily cannabis consumption is concentrated in adult populations, with noticeable variability that may reflect changing social patterns, policy shifts, or cultural attitudes over the study period.